Precise spiking motifs in neurobiological and neuromorphic data

### 2022-12-19 manuscript.tex
Update on Overleaf.

Another important algorithm, called SPOTDisClust, is based on the detection of structured temporal patterns~\citep{grossberger_unsupervised_2018}. They introduced an unsupervised method based for their detection from high-dimensional neural ensembles. The algorithm measures similarity between two ensemble spike patterns by determining the minimum transport cost of transforming their corresponding normalized cross-correlation matrices into each other. Many approaches to this problem are supervised. In other words, they take patterns occurring concurrently with a known event, such as the delivery of a stimulus for sensory neurons or the traversal of a running track for determining hippocampal place fields, as a ``template'' and then search for repetitions of the same template in spiking activity~\citep{nadasdy_replay_1999,lee_combinatorial_2004}. In SPOTDisClust, the learning is unsupervised. It uses the prior that there is only one spike per pattern. Using a so-called t-SNE projection (that allows to project this high-dimensional representation to a lower-dimension map) validated that this clustering method can retrieve all patterns from the data (see Figure~\ref{fig:SPOTDisClust}). The limits of this method are that it is computationally complex, block-based and strictly specialized for the task at hand. To overcome these difficulties, a novel method was recently developed~\citep{sotomayor-gomez_spikeship_2021}.
%   Davidson TJ, Kloosterman F, Wilson MA. Hippocampal replay of extended experience. Neuron. 2009;63(4):497--507. pmid:19709631
\add[JN]{Whether it is the distance between two given spike trains or the spike timings the complexity and the diversity of the methods used to measure them are witnesses of the growing interest of the integration of these measures in the understanding of the neural code. One of the steps to test their potential usefulness is then to scale these methods to larger amounts of data.}

@@

% \subsection{Extending the decoding at multiple time scales}
In the paper by~\citet{russo_cell_2017}, the authors present a unifying methodological and conceptual framework which detects assembly structure at many different time scales, levels of precision, and with arbitrary internal organization. It uses  sliding window as in~\citep{grun_unitary_2002} and the reliable and efficient analysis of an excess or deficiency of joint-spike events~\citep{pipa_neuroxidence_2008}. They extend the measure to multiple lags~\citep{torre_synchronous_2016}. The core measure is based on a non-stationarity corrected parametric statistical test for assessing the independence of pairs and an agglomerative, heuristic clustering algorithm for fusing significant pairs into higher-order assemblies.  To overcome the limits of models which either require spike times to be discretized, utilize a suboptimal least-squares criterion, or do not provide uncertainty estimates for model predictions or estimated parameters,~\citep{williams_point_2020} address each of these shortcomings by developing a point process model that characterizes fine-scale sequences at the level of individual spikes and represents sequence occurrences as a few marked events in continuous time. They also introduce learnable time-warping parameters to model sequences of varying duration, which have been experimentally observed in neural circuits, and demonstrate these advantages on experimental recordings from songbird higher vocal center and rodent hippocampus.
At a larger scale, in~\citep{van_kempen_top-down_2021}, it was shown that attentional information from V4 or arousal can change the timings of groups of events in V1. They develop a Hidden Markov Model for quantifying the transitions. In particular, they show that fluctuations in neural excitability are coordinated between visual areas with retinotopic precision. Top-down attention drives inter-areal coordination along the reverse cortical hierarchy, predicting better behavioral performance with increased coordination. Building such models for predicting changes of timings based on context, such as using a so-called change point model for blocked-based experimental protocols~\citep{pasturel_humans_2020}, could therefore provide useful prior information to enhance the decoding from neural activity.
% sparse in time and space : AL Barth and JF Poulet Trends in Neurosciences 35.6 (2012), pp.~345-355.  CC Petersen and S Crochet, Neuron 78.1 (2013), pp.~28-48.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{What biological mechanism could allow learning spiking motifs?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Despite the evidence for the effectiveness of precise spiking detection we presented above, doubts may remain as to the reliability of this learning mechanism and whether there is a real need for further research on this subject. The discovery of the existence of an equivalent biological mechanism in vertebrates as well as the demonstration of the importance of its role in various developed behaviors allow us to put these doubts to rest. In the following paragraphs we will successively describe the first biological observations of biological delay learning, identify myelinization as an important actor and finally study this phenomenon at the molecular level.

@@

However, it was not until the study of the interaural time difference (ITD) by \remove[Amélie]{gerstnerneuronal1996}~\citet{carr_circuit_1990} that it was discovered that this delay is not homogeneous for all neurons of the same type and species, but adapts according to their function. This ITD is a biological mechanism which allows for the azimuthal localization of sound by barn owls, by organically computing the difference in arrival time of a sound between their two ears. \add[Amélie]{It was first theorized in the Jeffress "ITD-versus-place model"}~\cite{jeffress}. \add[Amélie]{As hypothesized by this model, the \textit{nucleus laminaris} of the avian brain contains coincidence detectors and, associated with the \textit{nucleus magnocellularis} axons, forms circuits for processing ITD}~\cite{konishi}. According to~\citep{gerstner_neuronal_1996}, there is a true paradox in auditory neural systems since ``neural networks encode behaviorally relevant signals in the range of a few $\mu s$ with neurons that are at least one order of magnitude slower'', therefore necessarily need to play on the response time to do so. \add[Amélie]{This assertion confused the mean interspike interval, i.e. how often a neuron can fire, and the specific spike time, i.e. how precisely a spike can be generated. However, it has nevertheless contributed to the recognition of the importance of time in various biological mechanisms.} \remove[Amélie]{penacochlear2001 conducted an extended study on ITD in order to confirm this theory by identifying the coincidence detector involved, which the authors situated in the nuclear laminaris. In the mouse somatosensory cortex, camontiming2019 found a strong correlation between the delay of the mouse behavioral response and the timing of multiunit activity evoked by a trained whisker.} A first hypothesis suggested that the sound coincidence was detected using stereausis, i.e. the temporal disparity between left and right cochlear loci in the owl's brain. However, it was quickly set aside as the predictions did not match the measured disparities in the loci, and no variation was perceived in the nuclear laminaris for a similar sound intensity in both owl ears. The authors support a second hypothesis, that of different axonal delays in the ipsi- and contra-lateral cochlear nucleus magnocellularis.~\citet{seidl_mechanisms_2010} experimentally seconds this hypothesis of a ``coarse'' regulation of delay, as the authors conclude that regulations at different sites within individual axons of at least two parameters, namely the axon diameter and internode distances, might be responsible for precise adjustments of physiological delays thus allowing the ITD detection. \add[Amélie]{The authors also note that the barn owl's axons seem to change in length, thus implement a ``pure delay line''.} \add[Amélie]{The experiments described above thus conclude on the important role of physiological delays in the avian sound localisation behaviour. However, the relevance of precise timing in spikes is not limited to birds; for example, in the mouse somatosensory cortex,}~\citep{camon_timing_2019} \add[Amélie]{found a strong correlation between the delay of the mouse behavioral response and the timing of multiunit activity evoked by a trained whisker.} These experiments also confirmed previous studies stating that the conduction velocity of a spike in a neuron (in other words, its delay) depends strongly on the axon diameter~\citep{gasser_axon_1939} and the internode distance between Ranvier nodes~\citep{brill_conduction_1977}\change[Amélie]{--- instead of solely the axon's length as was previously thought.}{. This mechanism adds to the axonal length delay, which was previously thought sole influence on the conduction velocity due to its anatomical soundness --- as}~\citet{seidl_mechanisms_2010}\add[Amélie]{ experimentally demonstrated, this mechanism by itself is not sufficient to explain the biological functionality but should be added to the one of axonal length delay.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The importance of myelination}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% The importance of myelination
\citet{gasser_axon_1939} experimentally confirmed with homogeneously selected neurons that the axonal delay is positively proportional to \add[Amélie]{both the axon diameter by itself and the diameter of the axon wrapped in myelin}{the axon diameter, i.e. the amount of myelin wrapped around the axon}. Indeed, the oligodendrocytes, one of the many glial cells present in the \change[Antoine]{Vertebrates'}{vertebrates'} nervous system identified in 1924 by Pío del Rigo Hortega~\citep{perez-cerda_pio_2015}, produce thin protein sheets interspersed with lipid layers wrapped concentrically around the axon, called myelin~\citep{schmitt_ultrastructure_1939}. Myelinization consists in ``two motions: the wrapping of the leading edge of the inner tongue around the axon underneath the previously deposited membrane and the lateral extension of myelin membrane layers toward the nodal regions''~\citep{simons_oligodendrocytes_2016}. Multiple myelin regions can appear on one neuron axon and form the following subdomains: ``the internode (corresponding to the compacted region of myelin), the paranodes (where the outer loops of the myelin contact the axon), the juxtaparanode (the interface between the paranode and compact myelin, rich in potassium channels) and the node of Ranvier (the approx $1~\mu m$ gap between adjacent myelin internodes [allowing for] the saltatory conduction)''~\citep{duncan_neuron-oligodendrocyte_2021}. \change[Antoine]{In}{On} average, each oligodendrocyte produces 20 to 60 myelinization processes and each myelin sheath is 20 to 200 $\mu m$ long~\citep{simons_oligodendrocytes_2016}. This demonstrates an additional impact of myelinization on the conduction velocity, as the number of segments is positively correlated to the axonal delay~\citep{brill_conduction_1977}. Thus, in several occasions, myelin has been identified as \change[Antoine]{a main}{an important} actor in the regulation of conduction velocity in neurons, i.e. axonal delay regulation.

@@

% myelination for axonal delay, but what about dendritic delay ? 
The previous paragraphs present the biological mechanisms behind the axonal delay, regulated by myelination. However, the dendritic delay \add[Amélie]{as well as the axonal length delay (see }~\cite{seidl_mechanisms_2010}) seem to play an equally important role in the precise timing within a sequence of spikes. Dendritic delay is involved in the performance, structure and function of the nervous system, the modulation of spatio-temporal properties of pre- and post-synaptic activity patterns and the functional limitations of sensory feedback control efficiency~\citep{madadi_asl_dendritic_2018}. Its role has been specifically identified in the compensation of input asynchrony in the mammalian auditory brain stem~\citep{spencer_compensation_2018}. \cite{mel_dendrite} \add[Amélie]{highlights the dendrites' impact on neuronal plasticity, which is caused by the wide variation of numerous parameters: plasticity rules applied to different dendritic subtrees or dendritic subregions, local passive cable properties, distance travelled by remote dendritic inputs, branching structures, dendritic diameters, the relative timing of back-propagating somatic action potentials, etc. Dendritic spiking involvement in synaptic potentiation following active backpropagation into dendrites was experimentally uncovered using calcium imaging to highlight dendritic calcium entry allowing for long-tem potentiation}~\citep{golding}. ~\citet{branco_dendritic_2010} demonstrates the dendritic sensitivity to a sequence of synaptic activation in cortical pyramid neurons, encoded by ``both local dendritic calcium signals and somatic depolarization, leading to sequence-selective spike output''. The dendritic mechanism described can identify patterns delivered to a single dendrite or randomly distributed across the dendritic tree and relies on the dendritic calcium influx moderation by NMDA receptors.

@@

Spiking Neural Networks (SNNs)~\citep{maass_networks_1997} are natural candidates to use these precise temporal patterns in the brain. The approach which is currently most prominent in the SNN community is to use existing algorithms from machine learning and to adapt them to the specificity of spiking architectures~\citep{goltz_fast_2021}. One such example is to adapt the successes of deep learning algorithms and to transfer the back-propagation algorithm to SNNs, as it is the most widely used to tune the weights of a classical (non spiking) Neural Network. In a nutshell, it considers the system as implementing an input / output function and iteratively updates the weights according to the direction and magnitude of the error's gradient. Since the activation function of a spiking neuron is not differentiable, a recent popular approach consists in using a surrogate gradient~\citep{neftci_surrogate_2019} to ``cross-compile'' a classical Neural Network to a spiking architecture~\citep{rueckauer_conversion_2017}. This approach is quite successful, and SNNs reach in some case a similar performance as their non spiking equivalent, for instance on the N-MNIST dataset for categorizing digits in a stream of events~\citep{susi_nmnsd-spiking_2021}. The main reason for adopting this approach instead of classical architectures is the possibility to benefit from dedicated neuromorphic low-energy hardware. However, most biological neural systems are obviously more efficient than current state-of-the-art vision systems, both in terms of efficiency (accuracy), in speed (latency), and energy consumption. There is therefore an immense gap in the way we understand biology to translate it to the efficiency of SNNs.  To go beyond the state-of-the-art, we will focus here on one core computation of a spiking neuron, that is, is its ability to switch from the classical integrator mode (summing analog currents on its synapses) to a synchrony detector where it emits a spike whenever presynaptic spiking inputs are synchronized~\citep{abeles_role_1982,paugam-moisy_computing_2012}. In particular, we will explore different existing architectures which are able to overcome the diversity of input presynaptic patterns and learn to detect stable spiking motifs, that is, volleys of spikes which are stable up to a certain onset time (see Figure~\ref{fig:THC}). These models will be compared in light of neuroscientific and computational perspectives. %We review in this section theoretical and computational foundations of spiking motif detection. %In particular, our objective is to describe methods which fully exploit the capacity of spiking neurons to detect synchronous patterns.

>>>>

Another important algorithm, called SPOTDisClust, is based on the detection of structured temporal patterns~\citep{grossberger_unsupervised_2018}. They introduced an unsupervised method based for their detection from high-dimensional neural ensembles. The algorithm measures similarity between two ensemble spike patterns by determining the minimum transport cost of transforming their corresponding normalized cross-correlation matrices into each other. Many approaches to this problem are supervised. In other words, they take patterns occurring concurrently with a known event, such as the delivery of a stimulus for sensory neurons or the traversal of a running track for determining hippocampal place fields, as a ``template'' and then search for repetitions of the same template in spiking activity~\citep{nadasdy_replay_1999,lee_combinatorial_2004}. In SPOTDisClust, the learning is unsupervised. It uses the prior that there is only one spike per pattern. Using a so-called t-SNE projection (that allows to project this high-dimensional representation to a lower-dimension map) validated that this clustering method can retrieve all patterns from the data (see Figure~\ref{fig:SPOTDisClust}). The limits of this method are that it is computationally complex, block-based and strictly specialized for the task at hand. To overcome these difficulties, a novel method was recently developed~\citep{sotomayor-gomez_spikeship_2021}.
%   Davidson TJ, Kloosterman F, Wilson MA. Hippocampal replay of extended experience. Neuron. 2009;63(4):497--507. pmid:19709631
\add[JN]{Whether it is the distance between two given spike trains or a comparison of the spike timings, the complexity and the diversity of the methods used to measure them are witnesses of the growing interest of the integration of these measures in the understanding of the neural code. One of the steps to test their potential usefulness is then to scale these methods to larger amounts of data.}

@@

% \subsection{Extending the decoding at multiple time scales}
In the paper by~\citet{russo_cell_2017}, the authors present a unifying methodological and conceptual framework which detects assembly structure at many different time scales, levels of precision, and with arbitrary internal organization. It uses  sliding window as in~\citep{grun_unitary_2002} and the reliable and efficient analysis of an excess or deficiency of joint-spike events~\citep{pipa_neuroxidence_2008}. They extend the measure to multiple lags~\citep{torre_synchronous_2016}. The core measure is based on a non-stationarity corrected parametric statistical test for assessing the independence of pairs and an agglomerative, heuristic clustering algorithm for fusing significant pairs into higher-order assemblies.  To overcome the limits of models which either require spike times to be discretized, utilize a suboptimal least-squares criterion, or do not provide uncertainty estimates for model predictions or estimated parameters,~\citep{williams_point_2020} address each of these shortcomings by developing a point process model that characterizes fine-scale sequences at the level of individual spikes and represents sequence occurrences as a few marked events in continuous time. They also introduce learnable time-warping parameters to model sequences of varying duration, which have been experimentally observed in neural circuits, and demonstrate these advantages on experimental recordings from songbird higher vocal center and rodent hippocampus.
At a larger scale, in~\citep{van_kempen_top-down_2021}, it was shown that attentional information from V4 or arousal can change the timings of groups of events in V1. They develop a Hidden Markov Model for quantifying the transitions. In particular, they show that fluctuations in neural excitability are coordinated between visual areas with retinotopic precision. Top-down attention drives inter-areal coordination along the reverse cortical hierarchy, predicting better behavioral performance with increased coordination. Building such models for predicting changes of timings based on context, such as using a so-called change point model for blocked-based experimental protocols~\citep{pasturel_humans_2020}, could therefore provide useful prior information to enhance the decoding from neural activity.
% sparse in time and space : AL Barth and JF Poulet Trends in Neurosciences 35.6 (2012), pp.~345-355.  CC Petersen and S Crochet, Neuron 78.1 (2013), pp.~28-48.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{What biological mechanism could allow learning spiking motifs?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Despite the evidence for the effectiveness of precise spiking detection we presented above, doubts may remain as to the reliability of this learning mechanism and whether there is a real need for further research on this subject. The discovery of the existence of an equivalent biological mechanism in vertebrates as well as the demonstration of the importance of its role in various developed behaviors allow us to put these doubts to rest. In the following paragraphs we will successively describe the first biological observations of delay learning, identify myelinization as an important actor and finally study this phenomenon at the molecular level.

@@

However, it was not until the study of the interaural time difference (ITD) by \remove[Amélie]{gerstnerneuronal1996}~\citet{carr_circuit_1990} that it was discovered that this delay is not homogeneous for all neurons of the same type and species, but adapts according to their function. This ITD is a biological mechanism which allows for the azimuthal localization of sound by barn owls, by organically computing the difference in arrival time of a sound between their two ears. \add[Amélie]{It was first theorized in the Jeffress "ITD-versus-place model"}~\cite{jeffress_place_1948}. \add[Amélie]{As hypothesized by this model, the \textit{nucleus laminaris} of the avian brain contains coincidence detectors and, associated with the \textit{nucleus magnocellularis} axons, forms circuits for processing ITD}~\cite{konishi_coding_2003}. According to~\citep{gerstner_neuronal_1996}, there is a true paradox in auditory neural systems since ``neural networks encode behaviorally relevant signals in the range of a few $\mu s$ with neurons that are at least one order of magnitude slower'', therefore necessarily need to play on the response time to do so. \add[Amélie]{This assertion confused the mean interspike interval, i.e. how often a neuron can fire, and the specific spike time, i.e. how precisely a spike can be generated. However, it has nevertheless contributed to the recognition of the importance of time in various biological mechanisms.} \remove[Amélie]{penacochlear2001 conducted an extended study on ITD in order to confirm this theory by identifying the coincidence detector involved, which the authors situated in the nuclear laminaris. In the mouse somatosensory cortex, camontiming2019 found a strong correlation between the delay of the mouse behavioral response and the timing of multiunit activity evoked by a trained whisker.} A first hypothesis suggested that the sound coincidence was detected using stereausis, i.e. the temporal disparity between left and right cochlear loci in the owl's brain. However, it was quickly set aside as the predictions did not match the measured disparities in the loci, and no variation was perceived in the nuclear laminaris for a similar sound intensity in both owl ears. The authors support a second hypothesis, that of different axonal delays in the ipsi- and contra-lateral cochlear nucleus magnocellularis.~\citet{seidl_mechanisms_2010} experimentally seconds this hypothesis of a ``coarse'' regulation of delay, as the authors conclude that regulations at different sites within individual axons of at least two parameters, namely the axon diameter and internode distances, might be responsible for precise adjustments of physiological delays thus allowing the ITD detection. \add[Amélie]{The authors also note that the barn owl's axons seem to change in length, thus implementing a ``pure delay line''.} \add[Amélie]{The experiments described above thus conclude on the important role of physiological delays in the avian sound localisation behaviour. However, the relevance of precise timing in spikes is not limited to birds; for example, in the mouse somatosensory cortex,}~\cite{camon_timing_2019} \add[Amélie]{found a strong correlation between the delay of the mouse behavioral response and the timing of multiunit activity evoked by a trained whisker.} These experiments also confirmed previous studies stating that the conduction velocity of a spike in a neuron (in other words, its delay) depends strongly on the axon diameter~\citep{gasser_axon_1939} and the internode distance between Ranvier nodes~\citep{brill_conduction_1977}\change[Amélie]{--- instead of solely the axon's length as was previously thought.}{. This mechanism adds to the axonal length delay, which was previously thought sole influence on the conduction velocity due to its anatomical soundness --- as}~\citet{seidl_mechanisms_2010}\add[Amélie]{ experimentally demonstrated, this mechanism by itself is not sufficient to explain the biological functionality but should be added to the one of axonal length delay.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The importance of myelination}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% The importance of myelination
\citet{gasser_axon_1939} experimentally confirmed with homogeneously selected neurons that the axonal delay is positively proportional to \change[Amélie]{both the axon diameter by itself and the diameter of the axon wrapped in myelin}{the axon diameter, i.e. the amount of myelin wrapped around the axon}. Indeed, the oligodendrocytes, one of the many glial cells present in the \change[Antoine]{Vertebrates'}{vertebrates'} nervous system identified in 1924 by Pío del Rigo Hortega~\citep{perez-cerda_pio_2015}, produce thin protein sheets interspersed with lipid layers wrapped concentrically around the axon, called myelin~\citep{schmitt_ultrastructure_1939}. Myelinization consists in ``two motions: the wrapping of the leading edge of the inner tongue around the axon underneath the previously deposited membrane and the lateral extension of myelin membrane layers toward the nodal regions''~\citep{simons_oligodendrocytes_2016}. Multiple myelin regions can appear on one neuron axon and form the following subdomains: ``the internode (corresponding to the compacted region of myelin), the paranodes (where the outer loops of the myelin contact the axon), the juxtaparanode (the interface between the paranode and compact myelin, rich in potassium channels) and the node of Ranvier (the approx $1~\mu m$ gap between adjacent myelin internodes [allowing for] the saltatory conduction)''~\citep{duncan_neuron-oligodendrocyte_2021}. \change[Antoine]{In}{On} average, each oligodendrocyte produces 20 to 60 myelinization processes and each myelin sheath is 20 to 200 $\mu m$ long~\citep{simons_oligodendrocytes_2016}. This demonstrates an additional impact of myelinization on the conduction velocity, as the number of segments is positively correlated to the axonal delay~\citep{brill_conduction_1977}. Thus, in several occasions, myelin has been identified as \change[Antoine]{a main}{an important} actor in the regulation of conduction velocity in neurons, i.e. axonal delay regulation.

@@

% myelination for axonal delay, but what about dendritic delay ? 
The previous paragraphs present the biological mechanisms behind the axonal delay, regulated by myelination. However, the dendritic delay \add[Amélie]{as well as the axonal length delay (see }~\cite{seidl_mechanisms_2010}) seem to play an equally important role in the precise timing within a sequence of spikes. Dendritic delay is involved in the performance, structure and function of the nervous system, the modulation of spatio-temporal properties of pre- and post-synaptic activity patterns and the functional limitations of sensory feedback control efficiency~\citep{madadi_asl_dendritic_2018}. Its role has been specifically identified in the compensation of input asynchrony in the mammalian auditory brain stem~\citep{spencer_compensation_2018}. \cite{mel_synaptic_2017} \add[Amélie]{highlights the dendrites' impact on neuronal plasticity, which is caused by the wide variation of numerous parameters: plasticity rules applied to different dendritic subtrees or dendritic subregions, local passive cable properties, distance travelled by remote dendritic inputs, branching structures, dendritic diameters, the relative timing of back-propagating somatic action potentials, etc. Dendritic spiking involvement in synaptic potentiation following active backpropagation into dendrites was experimentally uncovered using calcium imaging to highlight dendritic calcium entry allowing for long-tem potentiation}~\citep{golding_dendritic_2002}. ~\citet{branco_dendritic_2010} demonstrates the dendritic sensitivity to a sequence of synaptic activation in cortical pyramid neurons, encoded by ``both local dendritic calcium signals and somatic depolarization, leading to sequence-selective spike output''. The dendritic mechanism described can identify patterns delivered to a single dendrite or randomly distributed across the dendritic tree and relies on the dendritic calcium influx moderation by NMDA receptors.

@@

Spiking Neural Networks (SNNs)~\citep{maass_networks_1997} are natural candidates to use these precise temporal patterns in the brain. The approach which is currently most prominent in the SNN community is to use existing algorithms from machine learning and to adapt them to the specificity of spiking architectures~\citep{goltz_fast_2021} One such example is to adapt the successes of deep learning algorithms and to transfer the back-propagation algorithm to SNNs, as it is the most widely used to tune the weights of a classical (non spiking) Neural Network. In a nutshell, it considers the system as implementing an input / output function and iteratively updates the weights according to the direction and magnitude of the error's gradient. \change[Antoine]{Since the activation function of a spiking neuron is not differentiable,}{In deep learning, gradient is computed on the activation function and since spikes are not differentiable,} a recent popular approach consists in using a surrogate gradient}~\citep{neftci_surrogate_2019} to ``cross-compile'' a classical Neural Network to a spiking architecture~\citep{rueckauer_conversion_2017}. SNNs reach in some case a similar performance as their non spiking equivalent, for instance on the N-MNIST dataset for categorizing digits in a stream of events~\citep{susi_nmnsd-spiking_2021}. \remove[Antoine]{The main reason for adopting this approach instead of classical architectures is the possibility to benefit from dedicated neuromorphic low-energy hardware.}{Even if improvements in terms of efficiency} However, most biological neural systems are obviously more efficient than current state-of-the-art vision systems, both in terms of efficiency (accuracy), in speed (latency), and energy consumption. There is therefore an immense gap in the way we understand biology to translate it to the efficiency of SNNs. To go beyond the state-of-the-art, we will focus here on one core computation of a spiking neuron, that is, is its ability to switch from the classical integrator mode (summing analog currents on its \change[Antoine]{synapses}{dendrites}) to a synchrony detector where it emits a spike whenever presynaptic spiking inputs are synchronized~\citep{abeles_role_1982,paugam-moisy_computing_2012}. In particular, we will explore different existing architectures which are able to overcome the diversity of input presynaptic patterns and learn to detect stable spiking motifs, that is, volleys of spikes which are stable up to a certain onset time (see Figure~\ref{fig:THC}). These models will be compared in light of neuroscientific and computational perspectives. %We review in this section theoretical and computational foundations of spiking motif detection. %In particular, our objective is to describe methods which fully exploit the capacity of spiking neurons to detect synchronous patterns.
